Carrier phase and amplitude estimation for phase shift keying using pilots and data

### 2013-11-16 response.tex

\begin{document}

\maketitle

\section*{Reviewer 1}

\begin{enumerate}
\item\textbf{Comment:} 
	This paper is concerned with carrier phase and amplitude estimation for PSK modulated signalling.
The considered least squares estimator seems to be only a minor extension of the one from [7]. Yet, more than 4 pages are dedicated to its derivation. Perhaps this section could be made more concise.
	\\\textbf{Response:}
We have shorteed Section~\zref{sec:least-squar-estim} but have left the description of the algorithm intact.  The primary reason for this is that much of the notation used during the analysis of the estimator is most naturally introduced during the description of the algorithm.  To qualify this for the reader we have included the following sentences at the beginning of Section~\zref{sec:least-squar-estim}.

``
In this section we derive Mackentun's algorithm to compute the least squares estimators of the carrier phase and amplitude~\cite{Mackenthun1994}.  Mackenthun specifically considered the non-data aided setting, so we modify the algorithm to include pilot symbols.  Although the modification is straightforward, much of the notation that we will use for analysing the estimator in Section~\ref{sec:stat-prop-least} is naturally introduced during the description of the algorithm.  For this reason, and so that our paper is self contained, we describe the algorithm in full.
''

%	We have shortened Sec.\ II to make it more concise and improve readability.
\item\label{R1it2}\textbf{Comment:} 
	As stated by the authors themselves, the key results and main contributions of the paper involve the asymptotic properties of the least squares estimator. As far as I can see the derivation seems correct and complete, yet it is mathematically very involved and rather difficult to follow.
	\\\textbf{Response:}	
	For improved readability of the manuscript, we have now moved all of the lemmas into the appendix. We have also added a discussion of the implications and the practical relevance of the theorems after the statement of Theorems~\zref{thm:consistency} and~\zref{thm:normality} in Section.~\zref{sec:stat-prop-least}.

%	\comment{Stress practical relevance by showing ASRP PER results. Mention trials and patents. Point out that knowledge of the estimator statistics makes it possible to assess the estimator performance for a given number of symbols/pilot symbols at a certain SNR without having to run time consuming simulations. Also point out that the results are valid for a wider range of noise distributions, whilst CRBs need to be derived for a specific noise distribution\dots}
\item
A few minor remarks are:
\begin{enumerate}
\item\label{R1it3a}\textbf{Comment:} 
	Estimation from data symbols only is referred to as noncoherent, I rather consider a detector with non-data-aided synchronization as coherent (since it uses estimates of the synchronization parameters to detect the data).
	\\\textbf{Response:}
	We agree with the reviewer and apologise for causing confusion with the use of the terms \emph{coherent} and \emph{noncoherent} in the original manuscript. In the updated manuscript, we refer to an estimator that does not use pilot symbols as \emph{non data-aided}.
%In the updated manuscript, we refer to an estimator that uses known pilot symbols as \emph{data-aided} and to one that relies on receiver decisions on the data symbols as \emph{decision-directed}.
\item\textbf{Comment:} 
	missing `,' after (5)
	\\\textbf{Response:} 
We thank the reviewer for spotting the missing comma and have added it.
\item\textbf{Comment:} 
	p.\ 19, at the bottom: sortindicIes (typo, delete 'i')
	\\\textbf{Response:} Thank you for pointing out this typographical error. We have corrected it.
\item\textbf{Comment:} 
	In Section VI, I would prefer a reference to the Lemma's in the Appendix at the start of the section.
	\\\textbf{Response:} We have added a reference to the Lemmas in the appendix at the beginning of the section. %Please note that Sec.\ VI of the original manuscript is now Sec.\ III.
\item\textbf{Comment:} 
	Section VIII, ``Theorem 2 does not model this behaviour\dots'' can you please further explain why? + Comparison with the Cramer-Rao bound results from the existing literature would be interesting.
	\\\textbf{Response:}
	We have removed the aforementioned sentence from the manuscript as we feel it has caused more confusion than provided insight. We hope the following explanation clears up any confusion it may have caused.  Theorem 2 shows that for $L\rightarrow\infty$, $\sqrt{L}\sfracpart{\theta_0-\hat{\theta}}_\pi$ is normally distributed with zero mean and finite variance.  The support of $\sqrt{L}\sfracpart{\theta_0-\hat{\theta}}_\pi$ is the interval $[-\sqrt{L}\pi, \sqrt{L}\pi)$, i.e.\ $(-\infty, \infty)$ as $L\rightarrow\infty$.  For finite $L$, however, $\sqrt{L}\sfracpart{\theta_0-\hat{\theta}}_\pi$ has finite support and thus, cannot be exactly normally distributed.  Nevertheless, provided the variance of the noise is sufficiently small (i.e.\ the SNR sufficiently high), the distribution of $\sqrt{L}\sfracpart{\theta_0-\hat{\theta}}_\pi$ will be concentrated in a finite interval much smaller than $[-\sqrt{L}\pi, \sqrt{L}\pi)$ and is very close to a normal distribution.  On the other hand, as the SNR decreases, $\sqrt{L}\sfracpart{\theta_0-\hat{\theta}}_\pi$ approaches a uniform distribution on $[-\sqrt{L}\pi, \sqrt{L}\pi)$.  Consequently, the variance of the phase estimator approaches that of the uniform distribution on $[-\pi, \pi)$.  The argument for the case without pilots is analogous.

	\comment{Add CRBs to plots.}
\end{enumerate}
\end{enumerate}

@@

\section*{Reviewer 2}
	General remarks
	Estimation of carrier phase and amplitude for $M$-ary PSK signals has been studied by the authors. Extending Mackenthun's algorithm to pilots, a least square (LS) method is developed, where major parts of the manuscript are dedicated to the statistical analysis of the LS approach.  Generally speaking it is to admit that both wording and writing follow the IEEE standards, although I would say that the list of References should be checked and updated in this respect.

@@

\item\textbf{Comment:} 
	Estimation of carrier phase and amplitude for $M$-ary PSK signals has been studied by the authors. Extending Mackenthun's algorithm to pilots, a least square (LS) method is developed, where major parts of the manuscript are dedicated to the statistical analysis of the LS approach.  Generally speaking it is to admit that both wording and writing follow the IEEE standards, although I would say that the list of References should be checked and updated in this respect.
	\\\textbf{Response:}
	We thank the reviewer for pointing us to the suggested references and have added them.

@@

\begin{enumerate}
\item\textbf{Comment:} 
	Paragraph before (13): \dots can be computed from its predecessor \dots
	\\\textbf{Response:} Thank you for pointing out this typographical error. We have corrected it.

@@

\item\textbf{Comment:} 
	Section III and VII: it would be helpful to provide $f(r, \phi)$
	\\\textbf{Response:}
	We thank the reviewer for pointing out the missing definition of $f(r, \phi)$ and have added it. %Please note that in response to comment \ref{R4it2e} of reviewer 4, Sec.\ III has been slightly shortened and has been moved to the appendix.

@@

\item\textbf{Comment:} 
	In the introductory section (paragraph before Eq.\ 3), the authors are talking about ``noncoherent detection where no pilot symbols exist''. In my mind this is somewhat confusing, because parameter estimation without knowing the underlying data is usually denoted as non-data aided (NDA) or decision-directed (DD), whereas when these data are known we have a data-aided (DA) situation [a]. On the other hand, coherent/noncoherent scenarios are given if the carrier phase is known/unknown to the receiver station.
	\\\textbf{Response:}
	Please refer to our response to comment \ref{R1it3a} of reviewer 1.

@@

\item\textbf{Comment:} 
	In a much more powerful approach to the topic addressed in the current paper, the pilot sequence would be employed for a coarse estimation/correction of the carrier phase of the payload data. In the sequel, by embedding carrier phase recovery into an iterative (turbo) decoding scheme [d, e] or any other error correcting method [f, g], the jitter performance is close to the CRLB for DA estimation even at very small SNRs, which are mainly determined by the convergence region of the (turbo) decoder. By taking into account the fact that modern and powerful communication systems are not conceivable without any FEC scheme, it is not clear to me why I should choose your algorithm in such a context.
	\\\textbf{Response:}
	We have added the following paragraph to the introduction describing the need for sufficiently accurate coarse estimates to initialise an iterative code-aided synchronisation stage. %citing [d, e].

``In the last decade, iterative methods for channel estimation that exploit the error correcting code used by the transmitter have appeared~\cite{Lottici2004,Noels2005,Herzet_turbo_synch_proc_IEEE_2007,Herzet_timingturbosynch_2007,Herzet_framework_turbo_sync_2007}.  These methods often go by the name of~\emph{turbo sychronisation}~\cite{Herzet_turbo_synch_proc_IEEE_2007}.  
These estimators typically apply the expectation maximisation (EM) algorithm under a Gaussian assumption regarding the noise sequence $\{w_i\}$.  A key problem is that the EM algorithm converges correctly only if initialised at some channel estimate sufficiently close to the true channel parameters.  Methods for efficiently obtaining a close initial channel estimate are still required and because of this the analysis of such methods remains of interest.  For the purpose of carrier phase and amplitude estimation, the least squares estimator we describe here is a good candidate as an initialiser for a turbo synchronisation proceedure.''

%We have also added a set of new simulation results, which show the packet error rate of an LDPC-coded satellite communication system to demonstrate the use and performance of the proposed estimator in a practical system.  \comment{Add details\dots}

@@

\section*{Reviewer 3}

\begin{enumerate}
\item
	\textbf{Comment:} 
The paper studies the estimation of amplitude and phase due to the distortion introduced by the channel in communication systems with $M$-PSK signals. The estimator is derived from the observation of $L$ noisy samples taken at a rate of one sample per symbol. Part of the $L$ observed samples corresponds to pilot transmitted symbols, that are already known a priori by the receiver, while the rest of samples are associated with data symbols which are unknown at the receiver. The focus of the article is to determine the amplitude and phase without considering the estimation of the transmitted data like this problem is usually treated in the literature. The work modifies the Mackenthun algorithm to include samples corresponding to pilot symbols while the Mackenthun paper calculates a least squares estimator for a similar problem but without pilot symbols.

The major contribution of this paper is the description of the asymptotic properties of the least squares estimator for the phase and amplitude. Although it seems technically correct, the authors do not make clear the relevance of their work to signal processing applications and to timely problems. They limit themselves in presenting several lemmas and proofs. Therefore, I cannot recommend this manuscript for publication in the IEEE Transactions on Signal Processing.
	\\\textbf{Response:}
We thank their reviewer for the comments. We have greatly expanded the introduction of the manuscript to make the practical and theoretical importance of our paper clear.
\end{enumerate}

@@

\section*{Reviewer 4}

\begin{enumerate}
\item\textbf{Comment:}
	This work addresses the problem of carrier phase and amplitude estimation for $M$-ary PSK signals. Based on a modification of the Mackhentum's algorithm by inserting pilots, the main contribution of this manuscript is in fact the study of the asymptotic behavior of the least squares estimator for carrier phase and amplitude. Although the derivations seems correct, it is not clear to the reader why this method should be applicable in a communications context.  A cconsiderable part of the manuscript is dedicated to proofs of Lemmas and Theorems, which makes the paper rather difficult to read. There are several methods in the litterature that proposes the use of pilots to estimate the carrier phase followed by efficient decoding such as turbo decoding. How does the proposed method compare with the existing uncoherent methods that do not require estimation of the amplitudes? Even if the data are coded, what are the gains in terms of data recovery of the proposed method and traditional solutions? In my opinion, the authors fail to convincing on this aspect and this is an important point.
	\\\textbf{Response:}
%	\comment{Stress practical relevance by showing ASRP PER results. In particular ST1, where the estimators are used to seed turbo sync. Mention trials and patents\dots}

%	\comment{Not clear what the reviewer means by ``existing uncoherent methods that do not require estimation of the amplitudes''.}

\item
Some additional comments:
\begin{enumerate}
\item\textbf{Comment:} 
	The authors should better motivate the interest in using their proposed method in actual communication systems. The lack of motivation for addressing the problem raised in the manuscript is a weak aspect of this work.
	\\\textbf{Response:}
%	\comment{Show PER ASRP results. Mention trials and patents\dots}
\item\textbf{Comment:} 
	The simulation results should bring comparisons with existing methods that use error correction and turbo decoding techniques, which have shown to perform reasonably well under adverse situations and low SNR levels.
	\\\textbf{Response:}
%	\comment{Show PER ASRP ST1 turbo sync results. Comparison with other estimators?}
\item\textbf{Comment:} 
	In Fig.\ 4, the influence of $P$ on the performance is not clear to me and does not seem to be in accordance with other results in the manuscript.
	\\\textbf{Response:} 
%	Please note that Fig.\ 4 in the original submission showed the variance of the amplitude estimator $\hat\rho$ (after bias removal), rather than the estimator MSE. This was done to validate the theoretical value of the asymptotic variance of $\hat\rho$ predicted by Theorem~2 against Monte-Carlo simulations. For consistency with the other figures and for easier interpretation, we have replaced the variance plot by a plot of the MSE of $\hat\rho$.
\item\textbf{Comment:} 
	It is important that the authors address an appropriate method for correcting the bias, as they mentioned in the text. A solution to provide the receiver with an estimate of $G(0)$ must be addressed in order to make this contribution applicable in a real communications context.
	\\\textbf{Response:} 
%	Biased estimators are quite common in the communications literature and in fact, are often used in practice. Prime examples are state-of-the-art signal-to-noise ratio (SNR) estimators for $M$-PSK modulated signals such as those discussed by Pauluzzi and Beaulieu \cite{Pauluzzi2000}. With the exception of the data-aided SNR estimators, all of the estimators in \cite{Pauluzzi2000} are biased and particularly at low SNR, the bias can be large. To the best of the authors' knowledge, the value of the bias of these estimators has only been determined by computer simulation and has not been analysed theoretically. In contrast, the asymptotic analysis of the amplitude estimator $\hat\rho$ presented in Theorem~1 of our manuscript results in a closed-form expression for the bias and thus, lays the foundation for improving the estimator by addressing the bias. However, removal of the bias is beyond the scope of this manuscript.
\item\textbf{Comment:} \label{R4it2e}
	Some sectioms can be shortened to make the work more readable and concise. This is the case of Section II for instance. Section III is not necessary. It is too short and the main concepts are well known. A reference to a textbook would be sufficient. Section VII is also too short and is not justified.
	\\\textbf{Response:}
%	We have shortened Sec.\ II to make it more concise and improve readability. 
\comment{BLERG}

The original Sec.\ VII has been removed from the revised manuscript.

\comment{BLERG}
%	Furthermore, we agree with the reviewer that circularly symmetric random variables are well known, but would like to point out that Sec.\ III specifically discusses random variables of the form $Re^{j\varPhi}=1{+}W$, where $W$ is circularly symmetric. The joint pdf of $R$ and $\varPhi$ is a crucial ingredient for the asymptotic analysis presented in the manuscript and therefore, we feel the section should be retained. However, for improved readability, Sec.\ III has been slightly shortened and has been moved to the appendix.
\end{enumerate}
\end{enumerate}

### 2013-11-16 paper.tex
added to paper

\begin{abstract}
We consider least squares estimators of carrier phase and amplitude from a noisy communications signal that contains both pilot signals, known to the receiver, and data signals, unknown to the receiver.  We focus on signaling constellations that have symbols evenly distributed on the complex unit circle, i.e., $M$-ary phase shift keying.  We show, under reasonably mild conditions on the distribution of the noise, that the least squares estimator of carrier phase is strongly consistent and asymptotically normally distributed.  However, the amplitude estimator is asymptotically biased and converges to a positive real number that is a function of the true carrier amplitude, the noise distribution and the size of the constellation.  
This appears to be the first time that the statistical properties of a non-data aided estimator for carrier amplitude have been analysed theoretically.
%Our theoretical results can also be applied to the case where no pilot symbols exist, i.e., noncoherent detection.  
The results of Monte Carlo simulations are provided and these agree with the theoretical results.   
\end{abstract}
\begin{IEEEkeywords}
Detection, phase shift keying, asymptotic statistics
\end{IEEEkeywords}

@@

We assume that %time offset estimation has been performed and that 
$L$ noisy $M$-PSK symbols are observed by the receiver,  %The received signal after matched filtering is,
\begin{equation}\label{eq:sigmod}
y_i = a_0 s_i + w_i, \qquad i \in P \cup D,
\end{equation}
where $w_i$ is noise and $a_0 = \rho_0 e^{j\theta_0}$ is a complex number representing both carrier phase $\theta_0$ and amplitude $\rho_0$ (by definition $\rho_0$ is a positive real number).  Our aim is to estimate $a_0$ from the noisy symbols $\{ y_i, i \in P \cup D \}$.  Complicating matters is that the data symbols $\{d_i, i \in D\}$ are not known to the receiver and must also be estimated.  Estimation problems of this type have undergone extensive prior study~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009,Divsalar1990}.  A practical approach is the least squares estimator, that is, the minimisers of the sum of squares function
\begin{equation}\label{eq:SSdefn}
\begin{split}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2  \\
&= \sum_{i \in P} \abs{ y_i - a p_i }^2 + \sum_{i \in D} \abs{ y_i - a d_i }^2,
\end{split}
\end{equation}
where $\abs{x}$ denotes the magnitude of the complex number $x$.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is white and Gaussian.  However, as we show, the estimator works well under less stringent assumptions.  %It is the least squares estimator that we primarily study in this paper.

% The existing literature~\cite{Mackenthun1994,Cowley_ref_sym_carr_1998,ViterbiViterbi_phase_est_1983,Sweldens2001,Wilson1989,Makrakis1990,Liu1991} mostly considers what is called \emph{noncoherent detection} (also know as \emph{blind}, \emph{non-data aided} or \emph{non-decision directed detection}) where no pilot symbols exist ($P = \emptyset$).  In the noncoherent setting \emph{differential encoding} is often used, and for this reason the estimation problem has been called \emph{multiple symbol differential detection}.  A popular approach is estimator of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  The idea is to `strip' the modulation from the received signal by taking $y_i / \abs{y_i}$ to the power of $M$.  A function $F: \reals \mapsto \reals$ is chosen and the estimator of the carrier phase $\theta_0$ is taken to be $\tfrac{1}{M}\angle{C}$ where $\angle$ denotes the complex argument and
% \begin{equation}\label{eq:viterbiviterbi}
% C = \frac{1}{L}\sum_{i \in P \cup D} F(\abs{y_i}) \big(\tfrac{y_i}{\abs{y_i}}\big)^M.
% \end{equation}
% Various choices for $F$ are suggested in~\cite{ViterbiViterbi_phase_est_1983} and a statistical analysis is presented.  A caveat of this estimator is that it is not obvious how pilot symbols should be included. %Techniques are available but they are somewhat ad hoc~\cite{Cowley_ref_sym_carr_1998}.  
% This problem does not occur with the least square estimator.

An important paper is by Mackenthun~\cite{Mackenthun1994} who described an algorithm to compute the least squares estimator requiring only $O(L \log L)$ arithmetic operations.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.  Both Mackenthun and Sweldens considered what is called \emph{non-data adied detection} (also know as \emph{blind}, \emph{noncoherent} or \emph{non-decision directed detection}) where no pilot symbols exist ($P = \emptyset$).  We show in Section~\ref{sec:least-squar-estim}~that Mackenthun's algorithm can be modified to include pilot symbols. Our model includes the non-data aided case by setting the number of pilot symbols to zero, that is, putting $P = \emptyset$.

Aside from this extention of Mackenthun's algorithm, the primary purpose of this paper is to analyse the least squares estimators of carrier phase and amplitude obtained by minimising~\eqref{eq:SSdefn}.  In the literature, analytic study of non-data aided estimators of carrier phase have primarily involved the derivation of Cramer-Rao lower bounds under the assumption that the noise $\{w_i\}$ is complex Gaussian distributed~\cite{Delmas_exact_crb_psk_2008,Steendam_lowsnr_crb_2001,Cowley_crbs_phase_freq_1996,DAndrea_modified_bounds_1994}.  The bounds themselves are often complicated to compute and analyse and for this reason it is common to make approximations, known as \emph{modified} Cramer-Rao bounds~\cite{DAndrea_modified_bounds_1994,Pollok_mckilliam_crb_cpm_2013}, or to analyse bounds that are valid asymptoticially as the noise variance increases to infinity or decreases to zero~\cite{Moeneclaey_limit_crbs_1998,Steendam_lowsnr_crb_2001,Delmas_exact_crb_psk_2008}.  To the authors knowledge, all of the existing bounds are applicable only in the fully non-data aided case or in the completely data aided case, that is, when either no pilot symbols exist, or when no data symbols exist.  Deriving bounds for the common practical scenario where there is a mixture of pilot and data symbols might be possible, but would further complicate already complicated bounds.

For the case of carrier amplitude the situations is even more complicated.  A simulation study conducted by Pauluzzi~\cite{Pauluzzi2000} suggests that known non-data aided estimators for carrier amplitude exhibit bias.  The bias is often small when the noise variance is small, but typically becomes significant as the noise variance increases.  Neverthess, Cramer-Rao bound have been derived for \emph{unbiased} estimators of carrier amplitude~\cite{Alagha_crb_snr_bpsk_qpsk_2001,Delmas_exact_crb_psk_2008}.  Unfortunately, these bounds may not be meaningful since known non-data aided amplitude estimators appear to be biased~\cite{Pauluzzi2000}.

In this paper we present an alternative approach to the study of estimators for carrier phase and amplitude.  Rather than compute bounds, we directly study the least squares estimators of carrier phase and amplitude asymptotically as the number of transmitted symbols $L\to \infty$.  We show, under comparatively mild conditions on the distribution of the noise, that the least squares estimator of carrier phase is strongly consistent (asymptotically unbiased) and asymptotically normally distributed.  However, the least squares amplitude estimator is asymptotically biased and converges to a positive real number that is a function of the true carrier amplitude, the noise distribution and the size of the constellation $M$.  To our knowledge, this is the first time an analytic description of the bias or variance of a non-data aided amplitude estimator has been obtained.  Our expressions for the asymptotic variance and bias are simple to compute.  %Our results are indirectly supported by the simulation study of Pauluzzi~\cite{Pauluzzi2000}.  We present our our set of simulations supporting our analytic results in Section~\ref{}.

@@

One criticism that might be directed at our asymptotic approach when compared to the Cramer-Rao bounds already developed in the literature is that our specific results apply only to the least square estimator.  One response to this criticism is that the least squares estimator should be a popular practical choice because it can be efficiently computed in $O(L\log L)$ operations and because it is the maximum likelihood estimator under the common assumption that the noise is white and Gaussian.  Perhaps an even more convinving response is that, while the specific results given here apply only to the least square estimtor, the techniques that we develop to derive these results are now able to be deployed for the analysis of other estimators.  For example, it is common in practical systems to place more importance on certain data symbols and use a small constellation size, say BPSK ($M=2$) for these, but to use a larger constellation size, say QPSK ($M=4$), for other less important symbols.  This is related to what is called \emph{unequal error protection} in the literature~\cite{Aydinlik_turbo_uep_2008,Sandberg_uep_ldpc_2010}.  It would be reasonably straigthforward to extend our results to allow data and pilot symbols with varying constellation size.  A more involved extension would be the inclusion of carrier frequency as a parameter to be estimated.  The techniques we present, combined with those from~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009,McKilliam_mean_dir_est_sq_arc_length2010}, would potentially allow this.

More ambitiously, our techniques might allow analysis of non-data aided channel estimators for modern communications systems such as mulitple anntannae systems, and systems employing orthogonal frequency division multiplexing~\cite{4138046,Heath99exploitinginput}.  We see the results presented here as a starting point for a potentially potent approach to the analysis of channel estimators that is complementary to the commonplace method of comparing results of Monte-Carlo simulations with Cramer-Rao lower bounds.

That this paper contains novel theoretical developments should not suggest to the reader that our results are not of immediate practical interest.  The design of our estimator was motivated by a project at the University of South Australia's Institute for Telecommunications Research to develop a narrowband multiuser communication system to collect sensor network data using a satellite in orbit.  This estimator has been implemented on a prototype that was successfully trialled from April to July of 2013 with signals from multiple low power ground-based transmitters being successfully received and decoded by both aircraft and satellites in orbit~\cite{ASRPpromovideo}.  Our theoretical analysis aided design of this system.  Based on our Theorem~\ref{thm:normality} in Section~\ref{sec:stat-prop-least} the performance of the least squares carrier phase and amplitude estimators can be accurately predicted without the need for time consuming Monte-Carlo simulations.  This provides easy to use and rigourously justified methods for choosing systems parameters such as the number and placement of pilot symbols and also provides useful information to other reciever components, such as channel tracking systems.

>>>>

\begin{abstract}
We consider least squares estimators of carrier phase and amplitude from a noisy communications signal that contains both pilot signals, known to the receiver, and data signals, unknown to the receiver.  We focus on signaling constellations that have symbols evenly distributed on the complex unit circle, i.e., $M$-ary phase shift keying.  We show, under reasonably mild conditions on the distribution of the noise, that the least squares estimator of carrier phase is strongly consistent and asymptotically normally distributed.  However, the amplitude estimator is asymptotically biased and converges to a positive real number that is a function of the true carrier amplitude, the noise distribution and the size of the constellation.  
This appears to be the first time that the statistical properties of a non data-aided estimator for carrier amplitude have been analysed theoretically.
%Our theoretical results can also be applied to the case where no pilot symbols exist, i.e., noncoherent detection.  
The results of Monte Carlo simulations are provided and these agree with the theoretical results.   
\end{abstract}
\begin{IEEEkeywords}
Detection, phase shift keying, asymptotic statistics
\end{IEEEkeywords}

@@

We assume that %time offset estimation has been performed and that 
$L$ noisy $M$-PSK symbols are observed by the receiver,  %The received signal after matched filtering is,
\begin{equation}\label{eq:sigmod}
y_i = a_0 s_i + w_i, \qquad i \in P \cup D,
\end{equation}
where $w_i$ is noise and $a_0 = \rho_0 e^{j\theta_0}$ is a complex number representing both carrier phase $\theta_0$ and amplitude $\rho_0$ (by definition $\rho_0$ is a positive real number).  Our aim is to estimate $a_0$ from the noisy symbols $\{ y_i, i \in P \cup D \}$.  Complicating matters is that the data symbols $\{d_i, i \in D\}$ are not known to the receiver and must also be estimated.  Estimation problems of this type have undergone extensive prior study~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009,Divsalar1990,580211,974597,Mengali_andre_synchro_book}.  A practical approach is the least squares estimator, that is, the minimisers of the sum of squares function
\begin{equation}\label{eq:SSdefn}
\begin{split}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2  \\
&= \sum_{i \in P} \abs{ y_i - a p_i }^2 + \sum_{i \in D} \abs{ y_i - a d_i }^2,
\end{split}
\end{equation}
where $\abs{x}$ denotes the magnitude of the complex number $x$.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is white and Gaussian.  However, as we show, the estimator works well under less stringent assumptions.  %It is the least squares estimator that we primarily study in this paper.

% The existing literature~\cite{Mackenthun1994,Cowley_ref_sym_carr_1998,ViterbiViterbi_phase_est_1983,Sweldens2001,Wilson1989,Makrakis1990,Liu1991} mostly considers what is called \emph{noncoherent detection} (also know as \emph{blind}, \emph{non data-aided} or \emph{non-decision directed detection}) where no pilot symbols exist ($P = \emptyset$).  In the noncoherent setting \emph{differential encoding} is often used, and for this reason the estimation problem has been called \emph{multiple symbol differential detection}.  A popular approach is estimator of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  The idea is to `strip' the modulation from the received signal by taking $y_i / \abs{y_i}$ to the power of $M$.  A function $F: \reals \mapsto \reals$ is chosen and the estimator of the carrier phase $\theta_0$ is taken to be $\tfrac{1}{M}\angle{C}$ where $\angle$ denotes the complex argument and
% \begin{equation}\label{eq:viterbiviterbi}
% C = \frac{1}{L}\sum_{i \in P \cup D} F(\abs{y_i}) \big(\tfrac{y_i}{\abs{y_i}}\big)^M.
% \end{equation}
% Various choices for $F$ are suggested in~\cite{ViterbiViterbi_phase_est_1983} and a statistical analysis is presented.  A caveat of this estimator is that it is not obvious how pilot symbols should be included. %Techniques are available but they are somewhat ad hoc~\cite{Cowley_ref_sym_carr_1998}.  
% This problem does not occur with the least square estimator.

An important paper is by Mackenthun~\cite{Mackenthun1994} who described an algorithm to compute the least squares estimator requiring only $O(L \log L)$ arithmetic operations.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.  Both Mackenthun and Sweldens considered what is called \emph{non-data adied detection} (also know as \emph{blind}, \emph{noncoherent} or \emph{non-decision directed detection}) where no pilot symbols exist ($P = \emptyset$).  We show in Section~\ref{sec:least-squar-estim}~that Mackenthun's algorithm can be modified to include pilot symbols. Our model includes the non data-aided case by setting the number of pilot symbols to zero, that is, putting $P = \emptyset$.

Aside from this extention of Mackenthun's algorithm, the primary purpose of this paper is to analyse the least squares estimators of carrier phase and amplitude obtained by minimising~\eqref{eq:SSdefn}.  In the literature, analytic study of non data-aided estimators of carrier phase have primarily involved the derivation of Cramer-Rao lower bounds under the assumption that the noise $\{w_i\}$ is complex Gaussian distributed~\cite{Delmas_exact_crb_psk_2008,950345,Steendam_lowsnr_crb_2001,Cowley_crbs_phase_freq_1996,DAndrea_modified_bounds_1994}.  The bounds themselves are often complicated to compute and analyse and for this reason it is common to make approximations, known as \emph{modified} Cramer-Rao bounds~\cite{DAndrea_modified_bounds_1994,Pollok_mckilliam_crb_cpm_2013}, or to analyse bounds that are valid asymptoticially as the noise variance increases to infinity or decreases to zero~\cite{Moeneclaey_limit_crbs_1998,Steendam_lowsnr_crb_2001,Delmas_exact_crb_psk_2008}.  To the authors knowledge, all of the existing bounds are applicable only in the fully non data-aided case or in the completely data aided case, that is, when either no pilot symbols exist, or when no data symbols exist.  Deriving bounds for the common practical scenario where there is a mixture of pilot and data symbols might be possible, but would further complicate already complicated bounds.

For the case of carrier amplitude the situations is even more complicated.  A simulation study conducted by Pauluzzi~\cite{Pauluzzi2000} suggests that known non data-aided estimators for carrier amplitude exhibit bias.  The bias is often small when the noise variance is small, but typically becomes significant as the noise variance increases.  Neverthess, Cramer-Rao bound have been derived for \emph{unbiased} estimators of carrier amplitude~\cite{Alagha_crb_snr_bpsk_qpsk_2001,Delmas_exact_crb_psk_2008}.  Unfortunately, these bounds may not be meaningful since known non data-aided amplitude estimators appear to be biased~\cite{Pauluzzi2000}.

In this paper we present an alternative approach to the study of estimators for carrier phase and amplitude.  Rather than compute bounds, we directly study the least squares estimators of carrier phase and amplitude asymptotically as the number of transmitted symbols $L\to \infty$.  We show, under comparatively mild conditions on the distribution of the noise, that the least squares estimator of carrier phase is strongly consistent (asymptotically unbiased) and asymptotically normally distributed.  However, the least squares amplitude estimator is asymptotically biased and converges to a positive real number that is a function of the true carrier amplitude, the noise distribution and the size of the constellation $M$.  To our knowledge, this is the first time an analytic description of the bias or variance of a non data-aided amplitude estimator has been obtained.  Our expressions for the asymptotic variance and bias are simple to compute.  %Our results are indirectly supported by the simulation study of Pauluzzi~\cite{Pauluzzi2000}.  We present our our set of simulations supporting our analytic results in Section~\ref{}.

@@

One criticism that might be directed at our asymptotic approach when compared to the Cramer-Rao bounds already developed in the literature is that our specific results apply only to the least square estimator.  One response to this criticism is that the least squares estimator should be a popular practical choice because it can be efficiently computed in $O(L\log L)$ operations and because it is the maximum likelihood estimator under the common assumption that the noise is white and Gaussian.  Perhaps an even more convinving response is that, while the specific results given here apply only to the least square estimator, the techniques that we develop to derive these results are now able to be deployed for the analysis of other estimators.  For example, in practical systems it is sometimes desirable to place more importance on certain data symbols and use a small constellation size, say BPSK ($M=2$) for these, but to use a larger constellation size, say QPSK ($M=4$), for other less important symbols.  This is related to what is called \emph{unequal error protection} in the literature~\cite{Aydinlik_turbo_uep_2008,Sandberg_uep_ldpc_2010}.  It would be reasonably straigthforward to extend our results to allow data and pilot symbols with varying constellation size.  A more involved extension would be the inclusion of carrier frequency as a parameter to be estimated.  The techniques we present, combined with those from~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009,McKilliam_mean_dir_est_sq_arc_length2010}, would potentially allow this.

More ambitiously, our techniques might allow analysis of non data-aided channel estimators for modern communications systems such as mulitple anntannae systems, systems employing orthogonal frequency division multiplexing, and systems that must widthstand rapidly time varying channels~\cite{4138046,Heath99exploitinginput,inthft:204506}.  We see the results presented here as a starting point for a potentially potent approach to the analysis of channel estimators that is complementary to the commonplace method of comparing results of Monte-Carlo simulations with Cramer-Rao lower bounds.

That this paper contains novel theoretical developments should does not imply that our results are of no immediate practical interest.  The design of our estimator was motivated by a project at the University of South Australia's Institute for Telecommunications Research to develop a narrowband multiuser communication system to collect sensor network data using a satellite in orbit.  This estimator has been implemented on a prototype that was successfully trialled from April to July of 2013 with signals from multiple low power ground-based transmitters being successfully received and decoded by both aircraft and satellites in orbit~\cite{ASRPpromovideo}.  Our theoretical analysis aided design of this system.  Based on our Theorem~\ref{thm:normality} in Section~\ref{sec:stat-prop-least} the performance of the least squares carrier phase and amplitude estimators can be accurately predicted without the need for time consuming Monte-Carlo simulations.  This provides easy to use and rigourously justified methods for choosing systems parameters such as the number and placement of pilot symbols and also provides useful information to other reciever components, such as channel tracking systems.
